---
title: "Lingcod epigenetic ageing v2 - FRS"
subtitle: "Fin ray ages - subset for otolith aged samples only"
author: "Alyssa"
date: "`r Sys.Date()`"
output:
  html_notebook:
    code_folding: hide
---

```{r setup, include=FALSE}
current_directory <- dirname(rstudioapi::getActiveDocumentContext()$path)
knitr::opts_knit$set(root.dir=normalizePath(paste0(current_directory)))
```

```{r load_libraries, include=FALSE}
# Load libraries 
library(caret)
library(glmnet)
library(ggpubr)
library(tidyverse)
```

```{r}
# Methylation matrix
methyl_data_unmod <- readRDS("dataFiles/06_methyl_data.rds")
```

```{r}
# Find line number where [Data] section starts
line_no <- grep("^\\[Data\\]", readLines("dataFiles/lingcod_sample_sheet_v1_21072025.csv"))

# Read the table starting from the line after [Data]
sample_sheet <- read.csv("dataFiles/lingcod_sample_sheet_v1_21072025.csv", skip = line_no, header = TRUE)

print("Samples in sample sheet but not methylation matrix: ")
setdiff(sample_sheet$Sample_Name, unique(methyl_data_unmod$sample))
```

```{r}
# Failed sample_ids
if (file.size("dataFiles/06_methyl_fails.txt") == 0) {
  print("No failed samples")
} else {
  methyl_fails <- read.table("dataFiles/06_methyl_fails.txt")
  failed_sample_ids <- sub("_.*", "", basename(methyl_fails$V1))
  print("Failed sample_ids (no sequencing data):")
  print(failed_sample_ids)
}
```

```{r}
# Import primer suite info
primer_suite_info_unmod <- read.csv("dataFiles/ophiodon_elongatus_primer_suite_mab21l2.csv")

primer_suite_info <- primer_suite_info_unmod %>%
  separate(region, into = c("chr", "range"), sep = ":", remove = FALSE) %>%
  separate(range, into = c("pos", "end_pos"), sep = "-", convert = TRUE)
```

```{r}
min_cov_threshold = 999
methyl_data_filtered <- methyl_data_unmod %>%
  mutate(coverage = meth_count + unmeth_count) %>%
  filter(coverage > min_cov_threshold)
```

```{r}
# Metadata
metadata_unmod <- read.csv("dataFiles/oelongatus_finray_ages.csv") 

metadata <- metadata_unmod %>%  
  rename_with(~ gsub("\\.\\.", ".", .) %>%     # Replace ".." with "."
                gsub("\\.$", "", .) %>%        # Remove trailing "."
                gsub("\\.", "_", .) %>%        # Replace remaining "." with "_"
                tolower()) %>%
  rename(age = fin_ray_age) %>%
  filter(!is.na(age)) # Remove samples with unknown age
```

```{r}
# Create coverage matrix
coverage_matrix <- methyl_data_filtered %>%
  select(site_id, sample, coverage) %>%
  pivot_wider(names_from = sample, values_from = coverage, values_fill = NA) %>%
  arrange(site_id) %>%
  column_to_rownames(var="site_id") %>%
  t(.)

# Create methylation matrix
methylation_matrix <- methyl_data_filtered %>%
  select(site_id, sample, meth_pct) %>%
  pivot_wider(names_from = sample, values_from = meth_pct, values_fill = NA) %>%
  arrange(site_id) %>%
  column_to_rownames(var="site_id") %>%
  t(.)

# # Export matrices
# write.csv(coverage_matrix, "dataFiles/07_coverage_matrix.csv", row.names=FALSE)
# write.csv(methylation_matrix, "dataFiles/07_methylation_matrix.csv", row.names=FALSE)
```

```{r fig.width=9, fig.height=12}
# Create heatmap
ggplot(coverage_matrix %>%
         as.data.frame(.) %>%
         # mutate(across(everything(), ~ replace_na(., 0))) %>%
         rownames_to_column(var="sample") %>%
         pivot_longer(-sample, names_to = "site_id", values_to = "methylation") ,
       aes(x = site_id, y = sample, fill = methylation)) +
  geom_tile() +
  scale_fill_viridis_c(option = "D", na.value = "white") +  # Use a color scale
  theme_minimal() +
  labs(x = "CpG site", y = "Sample", fill = "Number of reads") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size=8),
        axis.text.y = element_text(size=4))+
  ggtitle(paste0("Coverage matrix (white cells indicate < ", min_cov_threshold, " reads)"))
```

Beautiful even coverage - and lots of it!

```{r}
print("~~~ Coverage stats ~~~")
print(paste0("Minimum: ", min(methyl_data_filtered$coverage)))
print(paste0("Maximum: ", max(methyl_data_filtered$coverage)))
print(paste0("Mean: ", round(mean(methyl_data_filtered$coverage))))
print(paste0("Median: ", median(methyl_data_filtered$coverage)))
```

```{r fig.width=9, fig.height=12}
# Create heatmap
ggplot(methylation_matrix %>%
         as.data.frame(.) %>%
         rownames_to_column(var="sample") %>%
         pivot_longer(-sample, names_to = "site_id", values_to = "methylation") ,
       aes(x = site_id, y = sample, fill = methylation)) +
  geom_tile() +
  scale_fill_viridis_c(option = "magma", na.value = "white") +  # Use a color scale
  theme_minimal() +
  labs(x = "CpG site", y = "Sample", fill = "% Methylation") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size=8),
        axis.text.y = element_text(size=7))+
  ggtitle("DNA methylation matrix")
```

### Samples

```{r}
x <- length(row.names(methylation_matrix))
y <- length(metadata$sample_id)

print(paste0("The filtered DNAm matrix contains data for ", x, " of the total ", y, " fish samples."))
```

```{r}
print("Missing known-age samples: ")
missing_known_age_samples <- setdiff(metadata$sample_id, row.names(methylation_matrix))
missing_known_age_samples

pd_missing <- metadata_unmod %>%
  # filter(!is.na(Age)) %>%
  mutate(data = ifelse(Sample.ID %in% missing_known_age_samples, 
                       "NO DATA", 
                       "Yes")) %>%
  mutate(data = ifelse(Sample.ID == "", "Unknown", data))

# write.csv(pd_missing, "dataFiles/oelongatus_finray_age_missing.csv", row.names=FALSE)
```

```{r}
print("Sample size: ")
ss <- intersect(metadata$sample_id, row.names(methylation_matrix))
length(ss)
```

```{r}
# Randomly subset for 60 samples
# ss_sub <- sample(ss, size = 60, replace = FALSE)
```

```{r}
# Get otolith ones specifically
otolith_metadata_unmod <- read.csv("dataFiles/oelongatus_otolith_ages.csv") 

otolith_metadata <- otolith_metadata_unmod %>%  
  rename_with(~ gsub("\\.\\.", ".", .) %>%     # Replace ".." with "."
                gsub("\\.$", "", .) %>%        # Remove trailing "."
                gsub("\\.", "_", .) %>%        # Replace remaining "." with "_"
                tolower()) %>%
  rename(age = otolith_age) %>%
  filter(!is.na(age)) # Remove samples with unknown age

o_sub <- intersect(metadata$sample_id, otolith_metadata$sample_id)
```

```{r}
# Subset metadata for otolith ages
metadata <- metadata %>%
  filter(sample_id %in% o_sub)
```

```{r}
print("Updated sample size: ")
ss <- intersect(metadata$sample_id, row.names(methylation_matrix))
length(ss)
```

```{r}
# Model age

# Specify X and Y

# Ensure samples in Y but not X are removed
X <- as.data.frame(methylation_matrix) %>%
  rownames_to_column(., var = "sample_id") %>%
  filter(sample_id %in% metadata$sample_id) %>%
  column_to_rownames(., var="sample_id") %>%
  as.matrix(.)

# Filter for samples in X
Y <- metadata %>%
  filter(sample_id %in% rownames(X))
  
# Ensure X and Y have matching row names
Y <- Y %>%
  # Reorder 
  slice(match(rownames(X), sample_id))

# Split data into training (70%) and testing (30%)
set.seed(1234)
train_index <- createDataPartition(y=Y$age, p = 0.7, list = FALSE)
X_train <- X[train_index, ]
X_test <- X[-train_index, ]
Y_train <- Y[train_index, ]
Y_test <- Y[-train_index, ]

# Run cross-validated glmnet
cv_fit <- cv.glmnet(X_train, log(Y_train$age), alpha = 0.5)

# Run glmnet using lambda.min from cross-validation
glmnet_fit <- glmnet(X_train, log(Y_train$age), alpha = 0.5, lambda = cv_fit$lambda.1se)

# Predict on test set
Y <- Y %>%
  mutate(
    split = ifelse(sample_id %in% sample_id[train_index], " Train", "Test"),
    pred_age_log = as.numeric(predict(glmnet_fit, newx = X)),
    pred_age = exp(pred_age_log)
  ) %>%
  group_by(split) %>%
  mutate(rmse = sqrt(mean((age - pred_age)^2)),
         abs_error = abs(age - pred_age)) %>%
  ungroup()
```

```{r}
# Determine axis limits
axis_limits <- range(c(Y$age, Y$pred_age), na.rm = TRUE)

# Correlation labels
corr_labels <- Y %>%
  group_by(split) %>%
  summarise(cor = cor(age, pred_age), .groups = "drop") %>%
  mutate(
    label = paste0("r = ", round(cor, 2)),
    x = axis_limits[1] + 0.05 * diff(axis_limits),
    y = axis_limits[2] - 0.05 * diff(axis_limits)
  )

# Correlation plot
cplot <- ggplot(Y, aes(x = age, y = pred_age)) +
  geom_point(pch = 21, aes(fill = split), size = 2) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.8, colour = "black") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray") +
  geom_text(data = corr_labels, aes(x = x, y = y, label = label),
            inherit.aes = FALSE, check_overlap = TRUE, hjust = 0, size = 3.5) +
  # scale_fill_brewer(palette = "Set1", direction = -1) +
  scale_fill_manual(values=c("darkblue", "darkorange"))+
  facet_wrap(~split) +
  coord_cartesian(xlim = axis_limits, ylim = axis_limits) +
  labs(x = "Fin ray age", y = "Epigenetic age") +
  theme_bw() +
  theme(strip.background = element_blank())

# cplot
```

```{r}
# MAE labels
mae_labels <- Y %>%
  group_by(split) %>%
  summarise(mae = mean(abs_error), .groups = "drop") %>%
  mutate(
    label = paste0("MAE = ", round(mae, 2)),
    y = max(Y$abs_error, na.rm = TRUE) * 0.95,
    x = as.numeric(factor(split, levels = unique(Y$split)))
  )

# Absolute error plot
bplot <- ggplot(Y, aes(x = split, y = abs_error, fill = split)) +
  geom_boxplot(outlier.shape = NA, position = position_dodge(width = 0.8)) +
  geom_jitter(aes(color = split),
              position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.8),
              alpha = 0.5, size = 1.5) +
  geom_text(data = mae_labels, aes(x = x, y = y, label = label),
            inherit.aes = FALSE, size = 3.5, vjust = 0) +
  # scale_fill_brewer(palette = "Set1", direction = -1) +
  # scale_color_brewer(palette = "Set1", direction = -1) +
  scale_fill_manual(values=c("darkblue", "darkorange"))+
  scale_colour_manual(values=c("darkblue", "darkorange"))+
  labs(x = "Data split", y = "Absolute error (years)") +
  theme_bw()

# bplot
```

### Result

```{r fig.width=9, fig.height=3.5, message=FALSE}
ggarrange(cplot+theme(legend.position = "none"), 
          bplot+theme(legend.position = "none")+ggtitle(""), 
          ncol=2, widths=c(2,1))
```

*Note: A train/test split was used in this analysis to enable comparison with previous work. In future analyses, leave-one-out cross-validation (LOOCV) could easily be implemented to better align with common practices in the field.*

### Modelled sites/amplicons

```{r}
modelled_sites <- as.matrix(coef(cv_fit, s=cv_fit$lambda.1se))
modelled_sites <- as.matrix(modelled_sites[which(modelled_sites[,1] > 0 | modelled_sites[,1] < 0),])
```

```{r fig.width=9, fig.height=12}
# Create heatmap
ggplot(X %>%
         as.data.frame(.) %>%
         rownames_to_column(var="sample") %>%
         pivot_longer(-sample, names_to = "site_id", values_to = "methylation") %>%
         filter(site_id %in% row.names(modelled_sites)[-1]),
       aes(x = site_id, y = sample, fill = methylation)) +
  geom_tile() +
  scale_fill_viridis_c(option = "magma", na.value = "white") +  # Use a color scale
  theme_minimal() +
  labs(x = "CpG site", y = "Sample", fill = "% Methylation") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size=8),
        axis.text.y = element_text(size=7))+
  ggtitle("Predictive DNAm matrix (CpG sites selected by model)")
```

```{r}
# Amplicons that worked
adequate_sites <- data.frame(location=colnames(methylation_matrix)) %>%
  separate(location, into = c("chr", "range"), sep = ":", remove = FALSE) %>%
  separate(range, into = c("pos", "end_pos"), sep = "-", convert = TRUE)

# ID sites that shouldn't be there...
adequate_amplicons <- fuzzyjoin::genome_join(primer_suite_info %>% select(chr, pos, end_pos, region), 
                                       adequate_sites %>% select(chr, pos, end_pos), 
                                       by=c("chr", "pos", "end_pos")) %>%
  pull(region) %>%
  unique(.)

# Amplicons with data was used for modelling
predictive_sites <- as.data.frame(modelled_sites) %>%
  rownames_to_column(var="location") %>%
  filter(location != "(Intercept)") %>%
  separate(location, into = c("chr", "range"), sep = ":", remove = FALSE) %>%
  separate(range, into = c("pos", "end_pos"), sep = "-", convert = TRUE)

# ID sites that shouldn't be there...
predictive_amplicons <- fuzzyjoin::genome_join(primer_suite_info %>% select(chr, pos, end_pos, region), 
                                       predictive_sites %>% select(chr, pos, end_pos), 
                                       by=c("chr", "pos", "end_pos")) %>%
  pull(region) %>%
  unique(.)

x <- length(predictive_amplicons)
y <- length(adequate_amplicons)

n <- length(predictive_sites$location)
p <- length(adequate_sites$location)

print(paste0("To predict age, the model selected data from ", x, " amplicon (", n, " CpG sites) of the ", y, " total amplicons (", p, " CpG sites) that were adequately sequenced and modelled with elastic net"))
```

## Perform leave-one out cross validation (LOOCV) to replace test-train split

```{r message=FALSE}
# Model age

cntxt <- "CpG"

# Specify X and Y

# Ensure samples in Y but not X are removed
X <- as.data.frame(methylation_matrix) %>%
  rownames_to_column(., var = "sample_id") %>%
  filter(sample_id %in% metadata$sample_id) %>%
  column_to_rownames(., var="sample_id") %>%
  as.matrix(.)

# Filter for samples in X
Y <- metadata %>%
  filter(sample_id %in% rownames(X)) %>%
  # rename age
  mutate(age_years = age)
  
# Ensure X and Y have matching row names
Y <- Y %>%
  # Reorder 
  slice(match(rownames(X), sample_id))

# Perform LOOCV (using all data - not filtering for significant correlations first) ----
set.seed(1234)

# Specify alpha (0.5 for elastic net)
alpha_spec <- 0.5

pheno_loocv <- NULL
predictive_sites_loocv <- NULL

for (i in 1:nrow(Y)) {
  
  # Split data into testing and training folds ----
  
  # Use the sample_id_index to subset the training data
  Y_train <- Y[-i, ]
  
  # Use the inverse sample_id_index to subset the testing data
  Y_test <- Y[i, ]
  
  # Split into test and train
  X_train <- data.matrix(X[Y_train$sample_id, ])
  X_test <- data.matrix(X[Y_test$sample_id, , drop = FALSE]) ##**
  
  # Perform inner cross validation to optimise lambda (training data only) ----
  
  # Cross validate alpha and lambda
  cross_validation <- cv.glmnet(x=X_train, y=Y_train$age_years)
  
  # Get lambda 1se
  lambda_1se <- cross_validation$lambda.1se
  
  # Build the model (training data only) ----
  
  # Fit a glm with elastic net regularization to the training data
  model <- glmnet(X_train,
                  Y_train$age_years, 
                  family="gaussian",
                  lambda=lambda_1se,
                  alpha=alpha_spec)
  
  # Predict for training data (not used)
  Y_train <- Y_train %>%
    mutate(predicted_age_years = as.numeric(predict(model, 
                                         X_train, 
                                         type="response",
                                         s=lambda_1se))) %>%
    mutate(absolute_error = abs(age_years - predicted_age_years), 
           data_split = "Training data")
  
  # Predict for test data (n=1)
  Y_test <- Y_test %>%
    mutate(predicted_age_years = as.numeric(predict(model, 
                                         X_test, 
                                         type="response",
                                         s=lambda_1se))) %>%
    mutate(absolute_error = abs(age_years - predicted_age_years), 
           relative_error = absolute_error/age_years*100, 
           data_split = "Test data")
  
  # Combine
  pheno_loocv <- rbind(pheno_loocv, 
                       (full_join(Y_train, Y_test) %>%
                          mutate(iteration = i, # names(sample_id_indices[i]),
                                 lambda = lambda_1se))%>%
                         mutate(context = cntxt)) 
  
  # Extract the coefficients for the predictive sites
  predictive_sites <- data.frame(as.matrix(coef(model, s=lambda_1se))) %>%
    `colnames<-`("coefficient") %>%                # force a fixed name
    filter(coefficient != 0) %>%
    rownames_to_column(., var="id") %>%
    ## remove intercept
    filter(id != "(Intercept)") %>%
    ## reformat locations
    extract(id, into = c("seqid", "start", "end"),
            regex = "^(.+?):(\\d+)-(\\d+)$",
            remove = FALSE, convert = TRUE) %>%
    ## add identifiers
    mutate(cv_type = "LOOCV", 
           iteration = i) %>%
    mutate(context = toupper(cntxt))
  
  # Combine
  predictive_sites_loocv <- rbind(predictive_sites_loocv,
                                  predictive_sites) 
  
}
```

```{r}
## predicted ~ observed plot
loocv_test_results <- pheno_loocv %>%
  filter(data_split == "Test data") %>%
  group_by(context) %>%
  mutate(mae = round(mean(absolute_error), digits = 2),
         sd = round(sd(absolute_error), digits = 2),
         label = paste0("MAE: ", mae, " (\u00B1 ", sd, ")")) %>%
  mutate(mre = round(mean(relative_error), digits = 2),
         sd = round(sd(relative_error), digits = 2),
         labelr = paste0("MRE: ", mre, " (\u00B1 ", sd, ")")) %>%
  ungroup()

min_val <- min(c(loocv_test_results$age_years, loocv_test_results$predicted_age_years))
max_val <- max(c(loocv_test_results$age_years, loocv_test_results$predicted_age_years))

pc <- ggplot(loocv_test_results, aes(x=age_years, y=predicted_age_years))+
  geom_point(alpha = 0.5) + 
  geom_smooth(method = "lm") +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "grey")+
  labs(x = "Fin ray age (years)", 
       y = "Epigenetic age (years)") +
  lims(x=c(min_val, max_val), y=c(min_val, max_val))+
  # facet_grid(context~.)+
  theme_bw()+
  stat_cor(label.x.npc = "left", label.y.npc = "top", method = "pearson",
           aes(label = paste(after_stat(r.label))))+
  stat_regline_equation(label.x.npc = "left", label.y.npc = "top", vjust=4) +
  theme(strip.background = element_blank())

# pc
```

```{r}
## absolute error plot
ae <- ggplot(loocv_test_results, aes(x = age_years, y = absolute_error)) +
  geom_point(alpha=0.5)+
  geom_smooth(method = "lm", formula = y ~ poly(x, 2)) +
  labs(x = "Fin ray age (years)",
       y = "Absolute error (years)")+
  scale_y_continuous(breaks = scales::pretty_breaks())+
  ggpp::geom_text_npc(
    aes(npcx = 0.05,   # ≈ “near the start of the x-axis”
        npcy = 0.95,   # ≈ “near the top of the y-axis”
        label = label),
    hjust = 0, vjust = 1, check_overlap = TRUE)+
  theme_bw()+
  theme(strip.background = element_blank())

# ae
```

```{r}
## relative error plot
re <- ggplot(loocv_test_results, aes(x = age_years, y = relative_error)) +
  geom_point(alpha=0.5)+
  geom_smooth(method = "lm", formula = y ~ poly(x, 2)) +
  labs(x = "Fin ray age (years)",
       y = "Relative error (%)")+
  scale_y_continuous(breaks = scales::pretty_breaks())+
  ggpp::geom_text_npc(
    aes(npcx = 0.05,   # ≈ “near the start of the x-axis”
        npcy = 0.95,   # ≈ “near the top of the y-axis”
        label = labelr),
    hjust = 0, vjust = 1, check_overlap = TRUE)+
  theme_bw()+
  theme(strip.background = element_blank())

# re
```

```{r fig.width=10, fig.height=3.5, message=FALSE}
combined <- ggarrange(pc, ae, re, nrow=1)
combined

saveRDS(combined, file = "figuresTables/combined_finray_subset.rds")
```

```{r}
# Export data for comparisons
metadata_preds <- pheno_loocv %>%
  filter(data_split == "Test data") %>%
  select(sample_id, sex, age, predicted_age_years) %>%
  rename(fin_ray_age = age, 
         epigenetic_age = predicted_age_years) %>%
  mutate(epigenetic_age = round(epigenetic_age, digits=2))

write.csv(metadata_preds, "dataFiles/oelongatus_epiage_finraySubset.csv", row.names=FALSE)

write.csv(predictive_sites_loocv, "dataFiles/predsites_finraySubset", row.names=FALSE)
```
